spinifex: manual control of dynamic linear projections of high-dimensional data

### 2019-04-09 paper/spinifex-paper.tex
References to RJ format, touched up equations.

\abstract{%
The set of dynamic linear projections of multivariate data collectively known as \textbf{tours} provides an important tool for extending the dimensionality of visuals. The R package \textbf{tourr} offers a variety of path generators and geometric displays for conducting tours. This paper discusses an extension package, \textbf{spinifex}, that adds support for the path generation of manual tours and extends the display of tours to use with the animation packages, \textbf{plotly} and \textbf{gganimate}. Manual tours are used to explore the sensitivity of structure as the contributions of a manipulation variable are changed.
A recent paper \{Wang et al. (2018)\} visualizes the sensitivity of the hadronic experiments to nucleon structure. Sensitivity was characterized in non-linear 3D embeddings of the first 10 principal components. This research applies manual tours to this data showing that manual tours resolve information about cluster structure and identify outliers, information that is orthogonal to the original static viewing plane.
}

% Any extra LaTeX you need in the preamble

@@

A tour is a multivariate data analysis technique in which a sequence of linear (orthogonal) projections are viewed as an animation while the orientation of the projection basis is rotated across time. Each frame of the sequence corresponds to a small change in the projection for a smooth transition that perseveres continuity. This is a crucial tool extending the dimensionality of visualization which is a large boon as datasets and models exist in increasing cardinality. The package \textbf{tourr} (Wickham et al. 2011) builds a platform for generating tour paths and applying to a wide number of display geometrics in \textbf{base} graphics.

While there are numerous methods that generate tour paths, this research focuses on the manual tour. The manual tour was described in Cook and Buja (1997) and allows a user to control the projection coefficients of a selected variable in a 2D projection. The manipulation of these coefficients allows the analyst to explore how sensitive the projections structure is to these changes. This makes manual tours particularly useful once a feature of interest has been identified, for example, with the use of a guided tour (Hurley and Buja 1990). Rather than exploring the large phase space of the data, a guided tour selects a very specific path via projection pursuit(Diaconis and Freedman 1984), the optimization of an index function on the projection via a hill climbing algorithm. This allows guided tours to rapidly identify interesting projection features given the relatively large parameter-space. After a projection of interest is identified an analyst can then use the ``finer brush'' of the manual tour, by controlling the contributions of individual variables.

Spinifex utilizes two new animation packages, \textbf{plotly} (Sievert 2018) and \textbf{gganimate} (Pedersen and Robinson 2019), to display tours, manual or other saved tours. From a given projection, the user can choose which variable to control, and the animation sequence is generated to remove the variable from the projection, and then extend its contribution to be the sole variable in one direction. This allows the viewer to assess the change in structure induced in the projection by the variable's contribution.

@@

The steps are described in more detail below. The R functions used below mentioned briefly, but more complete code example can be found in section \ref{sec:usage}

@@

The flea data, originally from Lubischew (1962), made available in \textbf{tourr} is used to illustrate the algorithm. The data contains 74 observations across 6 variables, physical measurements of the flea beetles. Each observation belonging to one of three species.

The data is defined. A basis set (ideally orienting to an interesting feature) must be provided as an initial orientation. One way to identify a projection containing interesting features is to apply a guided tour (Cook, Swayne, and Buja 2007). In a guided tour the projection sequence is selected by optimizing an index function via hill-climbing on the projection space. In this case, the holes index is selected and applied to standardized flea data. The holes index is maximized when the projected observations are furthest from the center. Figure @ref\{fig:step0\} shows a locally optimized projection of the data. The left panel displays the reference axes of the projection basis, a visual indication of the magnitude and direction each variable contributed to the projection. The right panel shows the data as projected through the basis set described by the reference axes (left). Data points are colored and given shape according to the species (while the guided tour was unsupervised with this information).

(ref:step0-cap) Basis reference axes (left) and projected data (right) of standardized flea data. Data points color and shape are mapped to beetle species. Basis identified by a holes-index guided tour. The variables \texttt{aede2} and \texttt{tars1} contribute mostly orthogonal to the other variables.

@@

\caption[(ref:step0-cap)]{(ref:step0-cap)}(\#fig:step0)
\end{figure}
\end{Schunk}

Call \texttt{view\_basis()} on a basis to produce a \textbf{ggplot2} graphic similar to @ref(fig:step0). Projection space is always available for display via the matrix multiplication \(\textbf{X}_{[n,~p]} ~*~ \textbf{B}_{[p,~d]} ~=~ \textbf{P}_{[n,~d]}\).

@@

f_data  <- tourr::rescale(flea[,2:6])
f_path  <- save_history(f_data, guided_tour(holes()))
f_basis <- matrix(f_path[,, max(dim(f_path)[3])], ncol=2)
f_cat   <- factor(flea$species)

@@

In figure @ref(fig:step0) the contributions of the variables \texttt{tars1} and \texttt{aede2} are mostly orthogonal to the contributions of the other four variables. These two variables explain the variation of the data distinguishing the purple group from the rest of the sample. Select \texttt{aede2} as the manip var, the variable to be manipulated, as it has a larger contribution to the projection. The question that will be explored is how important the variable \texttt{aede2} is to the separation of the clusters.

@@

Adding this new dimension to our projection space allows for the coefficients of the manip var to be changed via rotation about the origin. For example, the ability to lift a piece of paper, rather than being constrained to a 2D plane. Orthonormalizing rescales the new vector while the projection down to 2D remains the original basis. Place the plane horizontally, with the new dimension extending vertically, with axes projecting back onto the reference axes. Figure @ref(fig:step2) illustrates this 3D manipulation space with the manip var highlighted (height of the other variables are not depicted.)

(ref:step2-cap) Manipulation space for controlling the contribution of \texttt{aede2} of standardized flea data. Basis selected by a holes-index optimized guided tour lies on the horizontal projection plane, shown in blue. The manip var axis, in red, extends into the vertical manipulation space, allowing the coefficients of the manip var to be changed by rotation around the origin.

@@

\caption[(ref:step2-cap)]{(ref:step2-cap)}(\#fig:step2)
\end{figure}
\end{Schunk}

The representation in figure @ref(fig:step2) can be produced by calling the function \texttt{view\_manip\_space()}.

\begin{Schunk}
\begin{Sinput}
view_manip_space(basis = f_basis, 
                 manip_var = 4, 
                 labels = colnames(f_data))
\end{Sinput}
\end{Schunk}

@@

then
\begin{align*}
  \textbf{RM}_{[p,~3,~i]}
  &= \textbf{M}_{[p,~3]} ~*~ \textbf{R}_{[3,~3,~i]} \\
  &= \begin{bmatrix}
    M_{1,~1} & M_{1,~2} & M_{1,~3} \\
    M_{2,~1} & M_{2,~2} & M_{2,~3} \\
    \vdots   & \vdots   \\
    M_{p,~1} & M_{p,~2} & M_{p,~3}
  \end{bmatrix}_{[p,~3]}
    ~*~
  \begin{bmatrix}
    c_\theta^2 c_\phi s_\theta^2 &
    -c_\theta s_\theta (1 - c_\phi) &
    -c_\theta s_\phi \\
    -c_\theta s_\theta (1 - c_\phi) &
    s_\theta^2 c_\phi + c_\theta^2 &
    -s_\theta s_\phi \\
    c_\theta s_\phi &
    s_\theta s_\phi &
    c_\phi
  \end{bmatrix}_{[3,~3,~i]}
\end{align*}

@@

Figure @ref(fig:step3) illustrates a sequence with 15 projected bases, showing the reference axes on top the corresponding projected data points on the below. As a result, we can see that changes in the manip var controlled the distance between the purple cluster and the remaining sample, \texttt{aede2} is crucial in distinguishing this species. Tours are typically viewed as an animation such a dynamic version of this tour can be viewed online at \url{https://nspyrison.netlify.com/thesis/flea_manualtour_mvar5/}. The page may take a moment to load.

(ref:step3-cap) Radial manual tour manipulating \texttt{aede2} of standardized flea data. The contributions increase from its initial contribution to a full contribution to the projection before decreasing to zero and then returning to its initial value. The change in the projected data shows that \texttt{aede2} is important for distinguishing the purple species. An animated version can be viewed at \url{https://nspyrison.netlify.com/thesis/flea_manualtour_mvar5/}.

@@

\caption[(ref:step3-cap)]{(ref:step3-cap)}(\#fig:step3)
\end{figure}
\end{Schunk}

@@

In a recent paper, Wang et al. (2018), the authors aggregate and visualize the sensitivity of hadronic experiments to nucleon structure. The authors introduce a new tool, PDFSense, to aid in the visualization of Parton distribution functions (PDF). The parameter-space of these experiments lies in 56 dimensions, \(\delta \in \mathbb{R}^{56}\), and are visualized as 3D subspaces of the 10 first principal components in linear (PCA) and non-linear (t-SNE) embeddings.

Using the same data, another study, Cook, Laa, and Valencia (2018), applies grand tours (Asimov 1985) to the same subspaces. Grand tours are dynamic subspace projections of high dimensional where frames and selected at random and linked with geodesically interpolation of the intermediate frames. Grand tours are able to better resolve the distribution shape of clusters, intra-cluster detail, and better outlier detection than the use of PDFSense \& TFEP (TensorFlow embedded projections). Before applying manual tours let's discuss the structure of the data.

@@

The \texttt{jet} cluster resides in a smaller dimensionality than the full set of experiments with four principal components explaining 95\% of the variation in the cluster (Cook, Laa, and Valencia 2018). The data within this 4D embedding is subset down to \texttt{ATLAS7old} and \texttt{ATLAS7new} to focus in on two groups with a reasonable number of observations that occupy different parts of the subspace. Radial manual tours controlling contributions from PC4 and PC3 are shown in figure @ref(fig:JetClusterGood) and figure @ref(fig:JetClusterBad) respectively. These components are selected to contrast the difference of information conveyed by touring different variables. Links to dynamic HTML5 animations controlling each of the four variables are also provided.

When manipulating PC4, there is clear difference in the parameter space spanned by the experiment types \texttt{ATLAS7new} and \texttt{ATLAS7old}. Yet, when PC3 is manipulated there is no indication that the diffent expiments probe different parameter space. Performing a radial manual tour on PC4 is more insightful than for PC3.

@@

(ref:JetClusterGood-cap) A radial manual tour of PC4 within the \texttt{jet} cluster. Colored by experiment type: \texttt{ATLAS7new} in green and \texttt{ATLAS7old} in orange. When PC4 fully/negligably contributes to the projection \texttt{ATLAS7new} (green) spans the same space as the orange points. During the intermediate frames the \texttt{ATLAS7new} is compressed in the direction radial to PC4. A dynamic version can be viewed at \url{https://nspyrison.netlify.com/thesis/jetcluster_manualtour_pc4/}.

@@

\caption[(ref:JetClusterGood-cap)]{(ref:JetClusterGood-cap)}(\#fig:JetClusterGood)
\end{figure}
\end{Schunk}

(ref:JetClusterBad-cap) A radial manual tour of PC4 within the \texttt{jet} cluster. Colored by experiment type: \texttt{ATLAS7new} in green and \texttt{ATLAS7old} in orange. Data from \texttt{ATLAS7new} (green) spans the mostly the same space as \texttt{ALTLAS7old} (orange) with no evident differnce in cluster structure across varying contributions of PC3. A dynamic version can be viewed at \url{https://nspyrison.netlify.com/thesis/jetcluster_manualtour_pc3/}.

@@

\caption[(ref:JetClusterBad-cap)]{(ref:JetClusterBad-cap)}(\#fig:JetClusterBad)
\end{figure}
\end{Schunk}

@@

A different space is used to explore the \texttt{DIS} cluster; specfically the first six pricipal components, which explain 48\% of the variation contained within the aggregated data(Cook, Laa, and Valencia 2018). Radial manual tours are performed on PC6 and PC2 in figure @ref(fig:DISclusterGood) and figure @ref(fig:DISclusterBad) respectively.

The selection of the manip variable is important as the manipulation spaces convey substantially different information. The manual tour of PC6 offers information about the dimensionality, shape, and orientations of the differnet experiment classes. whereas manipulating the contributions of PC2 only shows a subset of the dimensionality and shape information. Manipulating the contributions of PC6 turned out to be much more insightful than that of PC2. This result might seem counter-intuitive at first as PC2 should explain much more of the variation in the data. However, features and structure in the data regularly reside in finer detail that can be lost when looking only at static projections.

@@

(ref:DISclusterGood-cap) A radial manual tour manipulating the contribution of PC6 within the \texttt{DIS} cluster. Points are colored by experiment type: \texttt{DIS\ HERA1+2} in green, \texttt{dimuon\ SIDIS} in purple, and \texttt{charm\ SIDIS} in orange. The cluster \texttt{DIS\ HERA1+2} (green) is distributed in a cross shaped plane, \texttt{charm\ SIDIS} (orange) occupies the the center space of this cross, with the plane projecting into the field of view when the contribution of PC6 is max. Less evident is the linear \texttt{dimuon\ SIDIS} (purple) observations approaching the line of view for intermediate values of PC6. A dynamic version can be viewed at \url{https://nspyrison.netlify.com/thesis/discluster_manualtour_pc6/}.

@@

\caption[(ref:DISclusterGood-cap)]{(ref:DISclusterGood-cap)}(\#fig:DISclusterGood)
\end{figure}
\end{Schunk}

(ref:DISclusterBad-cap) A radial manual tour manipulating the contribution of PC2 within the \texttt{DIS} cluster. Points are colored by experiment type: \texttt{DIS\ HERA1+2} in green, \texttt{dimuon\ SIDIS} in purple, and \texttt{charm\ SIDIS} in orange. The plane of cross distributed \texttt{DIS\ HERA} data (green) and nearly orthogonal jet of \texttt{dimuon\ SIDIS} (purple) is present. This jet does entend more in the plane of view when the contribution of PC2 is full. However, less information about the shape of \texttt{DIS\ HERA} (green) and \texttt{charm\ SIDIS} (orange) is available. A dynamic version can be viewed at \url{https://nspyrison.netlify.com/thesis/discluster_manualtour_pc2/}.

@@

\caption[(ref:DISclusterBad-cap)]{(ref:DISclusterBad-cap)}(\#fig:DISclusterBad)
\end{figure}
\end{Schunk}

@@

Future research on the algorithm would include extending it for use in 3D projections. The addition of another dimension theoretically allows for improved perception. This could explore interactions in immersive virtual reality or mixed reality, which may further allow for a better perception of structure and aid in higher-dimensional function visualization. Functions with many parameters suffer from the same dimensionality problem as data while their possible values lie on a plane of values rather than discrete points. Occulation, or the closer surface blocking further surfaces, will likely be an issue that may be alleviated by the use of wire mesh, changing opacity, or looking at sections of the projections (Furnas and Buja 1994).

The \textbf{tourr} package provides many other geometric displays with the \texttt{tourr::display\_*()} family. These geometric options could be integrated into the \textbf{ggplot2} framework for display on \textbf{plotly} and \textbf{gganimate}. Additionally, the \textbf{animation} package Xie et al. (2018) could be implemented for another graphics framework. However, \textbf{animation} builds from \textbf{base} graphs while \textbf{spinifex} utilizes \textbf{ggplot2} graphics.

The Givens rotations and Householder reflections as outlined in Buja et al. (2005) could also be added. Currently, Gram-Schmidt is the only form of frame interpolation used (not used in manual tours). In a Givens rotation, the \(x\) and \(y\) components (for example \(\theta~= 0,~pi/2\)) of the in-plane rotation are calculated separately and would be applied sequentially to produce the radial rotation. Householder reflections define reflection axes to project points on to the axes and generate rotations.

Having a script only interaction with tours causes a significant barrier to entry. To a lesser extent, \textbf{plotly} offers some static interactions with the contained object, such as tooltips, brushing, and linking without communicating back to the R console. The development of a dynamic graphical user interface, perhaps with the use of a \textbf{shiny} (Chang et al. 2018) application, would mitigate the barrier to entry, allow for more rapid analysis, and offer an approachable demo tool. The user could easily switch between variables to control, adjust interpolation step angle, or flag/save specific frame basis sets.

@@

This article was created in R (R Core Team 2018), using \textbf{bookdown} (Xie 2016) and \textbf{rmarkdown} (Xie, Allaire, and Grolemund 2018), with code generating the examples inline. The source files for this article be found at \href{https://github.com/nspyrison/confirmation/}{github.com/nspyrison/confirmation/}. The source code for the \textbf{spinifex} package can be found at \href{https://github.com/nspyrison/spinifex/}{github.com/nspyrison/spinifex/}.

>>>>

\abstract{%
The set of dynamic linear projections of multivariate data collectively
known as \textbf{tours} provides an important tool for extending the
dimensionality of visuals. The R package \textbf{tourr} offers a variety
of path generators and geometric displays for conducting tours. This
paper discusses an extension package, \textbf{spinifex}, that adds
support for the path generation of manual tours and extends the display
of tours to use with the animation packages, \textbf{plotly} and
\textbf{gganimate}. Manual tours are used to explore the sensitivity of
structure as the contributions of a manipulation variable are changed. A
recent paper (Wang et al.
\protect\hyperlink{ref-wang_mapping_2018}{2018}) visualizes the
sensitivity of the hadronic experiments to nucleon structure.
Sensitivity was characterized in non-linear 3D embeddings of the first
10 principal components. This research applies manual tours to this data
showing that manual tours resolve information about cluster structure
and identify outliers, information that is orthogonal to the original
static viewing plane.
}

@@

A tour is a multivariate data analysis technique in which a sequence of
linear (orthogonal) projections are viewed as an animation while the
orientation of the projection basis is rotated across time. Each frame
of the sequence corresponds to a small change in the projection for a
smooth transition that perseveres continuity. This is a crucial tool
extending the dimensionality of visualization which is a large boon as
datasets and models exist in increasing cardinality. The package
\textbf{tourr} (Wickham et al.
\protect\hyperlink{ref-wickham_tourr_2011}{2011}) builds a platform for
generating tour paths and applying to a wide number of display
geometrics in \textbf{base} graphics.

While there are numerous methods that generate tour paths, this research
focuses on the manual tour. The manual tour was described in Cook and
Buja (\protect\hyperlink{ref-cook_manual_1997}{1997}) and allows a user
to control the projection coefficients of a selected variable in a 2D
projection. The manipulation of these coefficients allows the analyst to
explore how sensitive the projections structure is to these changes.
This makes manual tours particularly useful once a feature of interest
has been identified, for example, with the use of a guided tour (Hurley
and Buja \protect\hyperlink{ref-hurley_analyzing_1990}{1990}). Rather
than exploring the large phase space of the data, a guided tour selects
a very specific path via projection pursuit(Diaconis and Freedman
\protect\hyperlink{ref-diaconis_asymptotics_1984}{1984}), the
optimization of an index function on the projection via a hill climbing
algorithm. This allows guided tours to rapidly identify interesting
projection features given the relatively large parameter-space. After a
projection of interest is identified an analyst can then use the ``finer
brush'' of the manual tour, by controlling the contributions of
individual variables.

Spinifex utilizes two new animation packages, \textbf{plotly} (Sievert
\protect\hyperlink{ref-sievert_plotly_2018}{2018}) and
\textbf{gganimate} (Pedersen and Robinson
\protect\hyperlink{ref-pedersen_gganimate:_2019}{2019}), to display
tours, manual or other saved tours. From a given projection, the user
can choose which variable to control, and the animation sequence is
generated to remove the variable from the projection, and then extend
its contribution to be the sole variable in one direction. This allows
the viewer to assess the change in structure induced in the projection
by the variable's contribution.

@@

\hypertarget{fix-here}{%
\subsection{FIX here}\label{fix-here}}

The steps are described in more detail below. The R functions used below
mentioned briefly, but more complete code examples are given with

@@

The flea data, originally from Lubischew
(\protect\hyperlink{ref-lubischew_use_1962}{1962}), made available in
\textbf{tourr} is used to illustrate the algorithm. The data contains 74
observations across 6 variables, physical measurements of the flea
beetles. Each observation belonging to one of three species.

The data is defined. A basis set (ideally orienting to an interesting
feature) must be provided as an initial orientation. One way to identify
a projection containing interesting features is to apply a guided tour
(Cook, Swayne, and Buja
\protect\hyperlink{ref-cook_interactive_2007}{2007}). In a guided tour
the projection sequence is selected by optimizing an index function via
hill-climbing on the projection space. In this case, the holes index is
selected and applied to standardized flea data. The holes index is
maximized when the projected observations are furthest from the center.
Figure \ref{fig:step0} shows a locally optimized projection of the data.
The left panel displays the reference axes of the projection basis, a
visual indication of the magnitude and direction each variable
contributed to the projection. The right panel shows the data as
projected through the basis set described by the reference axes (left).
Data points are colored and given shape according to the species (while
the guided tour was unsupervised with this information).

@@

\caption[Basis reference axes (left) and projected data (right) of standardized flea data]{Basis reference axes (left) and projected data (right) of standardized flea data. Data points color and shape are mapped to beetle species. Basis identified by a holes-index guided tour. The variables aede2 and tars1 contribute mostly orthogonal to the other variables.}\label{fig:step0}
\end{figure}
\end{Schunk}

Call \texttt{view\_basis()} on a basis to produce a \textbf{ggplot2}
graphic similar to \textbackslash ref\{fig:step0). Projection space is
always available for display via the matrix multiplication
\(\textbf{X}_{[n,~p]} ~*~ \textbf{B}_{[p,~d]} ~=~ \textbf{P}_{[n,~d]}\).

@@

f_data  <- tourr::rescale(flea[, 1:6])
f_path  <- save_history(f_data, guided_tour(holes()))
f_basis <- matrix(f_path[,, max(dim(f_path)[3])], ncol=2)
f_cat   <- factor(flea$species)

@@

In figure \ref{fig:step0} the contributions of the variables
\texttt{tars1} and \texttt{aede2} are mostly orthogonal to the
contributions of the other four variables. These two variables explain
the variation of the data distinguishing the purple group from the rest
of the sample. Select \texttt{aede2} as the manip var, the variable to
be manipulated, as it has a larger contribution to the projection. The
question that will be explored is how important the variable
\texttt{aede2} is to the separation of the clusters.

@@

Adding this new dimension to our projection space allows for the
coefficients of the manip var to be changed via rotation about the
origin. For example, the ability to lift a piece of paper, rather than
being constrained to a 2D plane. Orthonormalizing rescales the new
vector while the projection down to 2D remains the original basis. Place
the plane horizontally, with the new dimension extending vertically,
with axes projecting back onto the reference axes. Figure
\ref{fig:step2} illustrates this 3D manipulation space with the manip
var highlighted (height of the other variables are not depicted.)

@@

\caption[Manipulation space for controlling the contribution of aede2 of standardized flea data]{Manipulation space for controlling the contribution of aede2 of standardized flea data. Basis selected by a holes-index optimized guided tour lies on the horizontal projection plane, shown in blue. The manip var axis, in red, extends into the vertical manipulation space, allowing the coefficients of the manip var to be changed by rotation around the origin.}\label{fig:step2}
\end{figure}
\end{Schunk}

The representation in figure \ref{fig:step2} can be produced by calling
the function \texttt{view\_manip\_space()}.

\begin{Schunk}
\begin{Sinput}
view_manip_space(basis = f_basis, 
                 manip_var = 5, 
                 labels = colnames(f_data))
\end{Sinput}
\end{Schunk}

@@

then \begin{align*}
  \textbf{RM}_{[p,~3,~i]}
  &= \textbf{M}_{[p,~3]} ~*~ \textbf{R}_{[3,~3,~i]} \\
  &= \textbf{M}_{[p,~3]}
    ~*~
  \begin{bmatrix}
    c_\theta^2 c_\phi s_\theta^2 &
    -c_\theta s_\theta (1 - c_\phi) &
    -c_\theta s_\phi \\
    -c_\theta s_\theta (1 - c_\phi) &
    s_\theta^2 c_\phi + c_\theta^2 &
    -s_\theta s_\phi \\
    c_\theta s_\phi &
    s_\theta s_\phi &
    c_\phi
  \end{bmatrix}_{[3,~3,~i]}
\end{align*}

@@

Figure \ref{fig:step3} illustrates a sequence with 15 projected bases,
showing the reference axes on top the corresponding projected data
points on the below. As a result, we can see that changes in the manip
var controlled the distance between the purple cluster and the remaining
sample, \texttt{aede2} is crucial in distinguishing this species. Tours
are typically viewed as an animation such a dynamic version of this tour
can be viewed online at
\url{https://nspyrison.netlify.com/thesis/flea_manualtour_mvar5/}. The
page may take a moment to load.

@@

\caption[Radial manual tour manipulating aede2 of standardized flea data]{Radial manual tour manipulating aede2 of standardized flea data.  The contributions increase from its initial contribution to a full contribution to the projection before decreasing to zero and then returning to its initial value. The change in the projected data shows that aede2 is important for distinguishing the purple species. An animated version can be viewed at  https://nspyrison.netlify.com/thesis/flea\_manualtour\_mvar5/.}\label{fig:step3}
\end{figure}
\end{Schunk}

@@

In a recent paper, Wang et al.
(\protect\hyperlink{ref-wang_mapping_2018}{2018}), the authors aggregate
and visualize the sensitivity of hadronic experiments to nucleon
structure. The authors introduce a new tool, PDFSense, to aid in the
visualization of Parton distribution functions (PDF). The
parameter-space of these experiments lies in 56 dimensions,
\(\delta \in \mathbb{R}^{56}\), and are visualized as 3D subspaces of
the 10 first principal components in linear (PCA) and non-linear (t-SNE)
embeddings.

Using the same data, another study, Cook, Laa, and Valencia
(\protect\hyperlink{ref-cook_dynamical_2018}{2018}), applies grand tours
(Asimov \protect\hyperlink{ref-asimov_grand_1985}{1985}) to the same
subspaces. Grand tours are dynamic subspace projections of high
dimensional where frames and selected at random and linked with
geodesically interpolation of the intermediate frames. Grand tours are
able to better resolve the distribution shape of clusters, intra-cluster
detail, and better outlier detection than the use of PDFSense \& TFEP
(TensorFlow embedded projections). Before applying manual tours let's
discuss the structure of the data.

@@

The \texttt{jet} cluster resides in a smaller dimensionality than the
full set of experiments with four principal components explaining 95\%
of the variation in the cluster (Cook, Laa, and Valencia
\protect\hyperlink{ref-cook_dynamical_2018}{2018}). The data within this
4D embedding is subset down to \texttt{ATLAS7old} and \texttt{ATLAS7new}
to focus in on two groups with a reasonable number of observations that
occupy different parts of the subspace. Radial manual tours controlling
contributions from PC4 and PC3 are shown in figure
\ref{fig:JetClusterGood} and figure \ref{fig:JetClusterBad}
respectively. These components are selected to contrast the difference
of information conveyed by touring different variables. Links to dynamic
HTML5 animations controlling each of the four variables are also
provided.

When manipulating PC4, there is a clear difference in the parameter
space spanned by the experiment types \texttt{ATLAS7new} and
\texttt{ATLAS7old}. Yet, when PC3 is manipulated there is no indication
that the different experiments probe different parameter space.
Performing a radial manual tour on PC4 is more insightful than for PC3.

@@

@@

\caption[A radial manual tour of PC4 within the jet cluster]{A radial manual tour of PC4 within the jet cluster. Colored by experiment type: ATLAS7new in green and ATLAS7old in orange. When PC4 fully/negligibly contributes to the projection ATLAS7new (green) spans the same space as the orange points. During the intermediate frames, the ATLAS7new is compressed in the direction radial to PC4. A dynamic version can be viewed at https://nspyrison.netlify.com/thesis/jetcluster\_manualtour\_pc4/.}\label{fig:JetClusterGood}
\end{figure}
\end{Schunk}

(ref:JetClusterBad-cap) A radial manual tour of PC4 within the
\texttt{jet} cluster. Colored by experiment type: \texttt{ATLAS7new} in
green and \texttt{ATLAS7old} in orange. Data from \texttt{ATLAS7new}
(green) spans mostly the same space as \texttt{ALTLAS7old} (orange) with
no evident difference in cluster structure across varying contributions
of PC3. A dynamic version can be viewed at
\url{https://nspyrison.netlify.com/thesis/jetcluster_manualtour_pc3/}.

@@

\caption[(ref:JetClusterBad-cap)]{(ref:JetClusterBad-cap)}\label{fig:JetClusterBad}
\end{figure}
\end{Schunk}

@@

A different space is used to explore the \texttt{DIS} cluster;
specifically the first six principal components, which explain 48\% of
the variation contained within the aggregated data(Cook, Laa, and
Valencia \protect\hyperlink{ref-cook_dynamical_2018}{2018}). Radial
manual tours are performed on PC6 and PC2 in figure
\ref{fig:DISclusterGood} and figure \ref{fig:DISclusterBad}
respectively.

The selection of the manip variable is important as the manipulation
spaces convey substantially different information. The manual tour of
PC6 offers information about the dimensionality, shape, and orientations
of the different experiment classes. whereas manipulating the
contributions of PC2 only shows a subset of the dimensionality and shape
information. Manipulating the contributions of PC6 turned out to be much
more insightful than that of PC2. This result might seem
counter-intuitive at first as PC2 should explain much more of the
variation in the data. However, features and structure in the data
regularly reside in finer detail that can be lost when looking only at
static projections.

@@

@@

\caption[A radial manual tour manipulating the contribution of  PC6 within the DIS cluster]{A radial manual tour manipulating the contribution of  PC6 within the DIS cluster. Points are colored by experiment type: DIS HERA1+2 in green, dimuon SIDIS in purple, and charm SIDIS in orange. The cluster DIS HERA1+2 (green) is distributed in a cross-shaped plane, charm SIDIS (orange) occupies the center space of this cross, with the plane projecting into the field of view when the contribution of PC6 is max. Less evident is the linear dimuon SIDIS (purple) observations approaching the line of view for intermediate values of PC6. A dynamic version can be viewed at https://nspyrison.netlify.com/thesis/discluster\_manualtour\_pc6/.}\label{fig:DISclusterGood}
\end{figure}
\end{Schunk}

@@

\caption[A radial manual tour manipulating the contribution of PC2 within the DIS cluster]{A radial manual tour manipulating the contribution of PC2 within the DIS cluster. Points are colored by experiment type: DIS HERA1+2 in green, dimuon SIDIS in purple, and charm SIDIS in orange. The plane of cross distributed DIS HERA data (green) and a nearly orthogonal jet of dimuon SIDIS (purple) is present. This jet does extend more in the plane of view when the contribution of PC2 is full, giving insight to its orientation. However, less information about the shape of DIS HERA (green) and charm SIDIS (orange) is available. A dynamic version can be viewed at https://nspyrison.netlify.com/thesis/discluster\_manualtour\_pc2/.}\label{fig:DISclusterBad}
\end{figure}
\end{Schunk}

@@

Future research on the algorithm would include extending it for use in
3D projections. The addition of another dimension theoretically allows
for improved perception. This could explore interactions in immersive
virtual reality or mixed reality, which may further allow for a better
perception of structure and aid in higher-dimensional function
visualization. Functions with many parameters suffer from the same
dimensionality problem as data while their possible values lie on a
plane of values rather than discrete points. Occulation, or the closer
surface blocking further surfaces, will likely be an issue that may be
alleviated by the use of wire mesh, changing opacity, or looking at
sections of the projections (Furnas and Buja
\protect\hyperlink{ref-furnas_prosection_1994}{1994}).

The \textbf{tourr} package provides many other geometric displays with
the \texttt{tourr::display\_*()} family. These geometric options could
be integrated into the \textbf{ggplot2} framework for display on
\textbf{plotly} and \textbf{gganimate}. Additionally, the
\textbf{animation} package Xie et al.
(\protect\hyperlink{ref-xie_animation:_2018}{2018}) could be implemented
for another graphics framework. However, \textbf{animation} builds from
\textbf{base} graphs while \textbf{spinifex} utilizes \textbf{ggplot2}
graphics.

The Givens rotations and Householder reflections as outlined in Buja et
al. (\protect\hyperlink{ref-buja_computational_2005}{2005}) could also
be added. Currently, Gram-Schmidt is the only form of frame
interpolation used (not used in manual tours). In a Givens rotation, the
\(x\) and \(y\) components (for example \(\theta~= 0,~pi/2\)) of the
in-plane rotation are calculated separately and would be applied
sequentially to produce the radial rotation. Householder reflections
define reflection axes to project points on to the axes and generate
rotations.

Having a script only interaction with tours causes a significant barrier
to entry. To a lesser extent, \textbf{plotly} offers some static
interactions with the contained object, such as tooltips, brushing, and
linking without communicating back to the R console. The development of
a dynamic graphical user interface, perhaps with the use of a
\textbf{shiny} (Chang et al.
\protect\hyperlink{ref-chang_shiny:_2018}{2018}) application, would
mitigate the barrier to entry, allow for more rapid analysis, and offer
an approachable demo tool. The user could easily switch between
variables to control, adjust interpolation step angle, or flag/save
specific frame basis sets.

@@

This article was created in R (R Core Team
\protect\hyperlink{ref-r_core_team_r:_2018}{2018}), using
\textbf{bookdown} (Xie \protect\hyperlink{ref-xie_bookdown:_2016}{2016})
and \textbf{rmarkdown} (Xie, Allaire, and Grolemund
\protect\hyperlink{ref-xie_r_2018}{2018}), with code generating the
examples inline. The source files for this article be found at
\href{https://github.com/nspyrison/confirmation/}{github.com/nspyrison/confirmation/}.
The source code for the \textbf{spinifex} package can be found at
\href{https://github.com/nspyrison/spinifex/}{github.com/nspyrison/spinifex/}.

\hypertarget{bibliography}{%
\subsection*{Bibliography}\label{bibliography}}
\addcontentsline{toc}{subsection}{Bibliography}
